# R basics

Before you start, download and install **both** R ([download  R](https://cran.rstudio.com)) **and** RStudio ([download RStudio](https://rstudio.com/products/rstudio/download/#download)) if you haven't yet.

I recommend working in Rstudio for its user-friendly interface and useful additional functionalities.

Start RStudio and open a new script: `File` >` New File` > `R Script`. The **R script window** is what you'll use to **write and save** your code. Write in your new script `"Welcome!"` and run the line using `Ctrl`+`Enter` (Windows/Linux) or `Cmd`+`Return` (Mac). The **output** is displayed in the **console window** (by default, it's the bottom left window) and should be `[1] "Welcome!"` in this case.

Just like other coding and programming languages, R is prone to syntax errors, e.g. typos, forgetting a quote, bracket or comma. Such mistakes will break your code and throw an error. Syntax errors are the most common type of error, so always double-check your code whenever R is unhappy.

## Data types

R treats all data as objects (think objects as nouns and adjectives). The most basic type of object in R is called a vector. E.g. `24` and `"hi"` are two different vectors of one element. Different vectors have different data types.

There're six data types in R. Here, we'll focus on the four most common ones: `character`, `logical`, `double` and `integer`.

```{r}
#| label: data-types-tbl
#| echo: false

tibble::tribble(
  ~Value, ~Type, ~Abbrevation,
  "-1, 0, 2.38", "double", "dbl",
  "-1L, 0L, 2L", "integer", "int",
  '"Hello world!"', "character (enclosed in double or single quotes)", "chr",
  "TRUE, FALSE", "logical", "lgl"
) |>
  knitr::kable()
```

> [!NOTE]
> `double` and `integer` types belong to the `numeric` class type.

You can perform mathematical operations on `numeric` and `logical` vectors but not on `character` vectors.

```{r}
#| label: example-data-types
#| error: true

1 + 1

TRUE + TRUE

"1" + "1"
```

## Vectors

You can combine multiple vectors into one using the `c()` [function](#functions). In this case, the vector is a sequence of **elements** of the **same type** (also known as atomic vector).

```{r}
#| label: vec
#| results: "hide"

# double-precision vector
c(1, -2, 430.78)

# integer vector
c(1L, -2L, 430L)

# character vector
c("foo", "bar")

# logicial vector
c(TRUE, FALSE, FALSE)
```

`c()` automatically converts each element to the same type.

```{r}
#| label: vec-different-types

# despite having a single character element,
# the entire set is converted to the character type

c(1, "Hi", TRUE)
```

You can combine vectors of different type using `list()`:

```{r}
#| label: list
#| collpase: true
list(1, "Hi", TRUE)
```

## Variables

Variables provide a storage space to R objects. You can store all type of data in a variable, including chunk of code or another variable.

To assign a name to your object, use the assignment operator `<-` (keyboard shorcut: `Alt`+`-` (Windows & Linux) or	`Option`+`-` (Mac)).

```{r}
#| label: var-syntax
#| eval: false
my_variable <- my_object
```

> [!NOTE]
> The name of the variable can be a combination of `letters`, `digits`, `.` and `_`. You're free to give your variable the name you want as long as it **starts with** a `letter` or `.`. Choose a name that is both **concise** and **explicit**.

To print what is stored in a variable, simply call that variable.

```{r}
#| label: example-vars

hello_wold <- "Hello world!"
hello_wold

x <- 2
x

y <- 4
z <- x * y
z
```

If two or more variables in the same environment have the same name, R will return the value of the variable defined the last.

```{r}
#| label: example-vars-same-name

value <- 3
value <- 4
value
```

> [!NOTE]
> R is case sensitive. This means that lower and upper case letters are interpreted as different characters. E.g. `foo` is **not** the same as `Foo`.

## Functions

A function is what you use to perform an action on your objects. Similarly to objects, which can be thought as nouns and adjectives, functions can be thought as verbs.

A function can take one, several or no arguments at all.

```{r}
#| label: fn-syntax
#| eval: false

# a function with multiple arguments
function_a(arg_1, arg_2)

# a function with no arguments
function_b()
```

> [!NOTE]
> Most arguments are optional. Arguments are separated by a `,`.

Let's sum our previous `x` and `y` objects using the `sum()` function:

```{r}
#| label: example-fn
sum(x, y)
```

You can also create your own functions:

```{r}
#| label: function

# function to convert meters to feet
m_to_feet <- function(value) {
  value / .3048
}

m_to_feet(100)
```

Values are assigned to argument names in the order they're defined. Writing the name of each argument isn't mandatory but can lead to unwanted results.

```{r}
#| label: arg-order

divide <- function(dividend, divisor) {
  dividend / divisor
}

divide(x, y)

divide(y, x)

divide(divisor = y, dividend = x)
```

It's common to omit the first two argument names of frequently used functions but I highly recommend writing the name of any subsequent argument in functions that take more than two arguments.

## Relational and logical operators

You can use relational and logical operators to compare values and perform boolean operations. These operators yield a boolean value, that is either `TRUE` or `FALSE`.

Below is a list of R's most common relational and logical operators.

```{r}
#| label: logical-operators-tbl
#| echo: false

tibble::tribble(
  ~Operator, ~Description,
  "!", "not",
  "==", "equal to",
  "!=", "not equal to",
  "<", "less than",
  ">", "greater than",
  "<=", "less than or equal to",
  ">=", "greater than or equal to",
  "%in%", "check if one or more elements exist in a vector",
  "&", "and",
  "|", "or"
) |>
  knitr::kable()
```

Here're some examples to illustrate how these operators work:

```{r}
#| label: example-logical-operators

"foo" == "foo"

2 != 3.5

4 < -2

c("Hello", "Bye") %in% c("Hello", "world", "!")

!(2 %in% c(3, 6, 2))

TRUE & FALSE

TRUE | FALSE
```

## Comments

You can write comments in your script to add details and information to your code, or to stop a line from running. Comments are muted, meaning they aren't interpreted by the console.

To add a comment, simply insert a `#` before the statement you want to comment.

```{r}
#| label: example-comments

# comments aren't interpreted by the console
x <- 3

# x <- 4

x
```
