# Exercise

Using the `penguins` dataset, look at the relationship between bill area and body mass in adelie and chinstrap penguins.

In a new column, compute the bill area for each individual. Plot the bill area against the body mass by species group using `ggplot()` and `geom_point()`.

For the purpose of this exercise, we'll make the assumption that the bill is flat and that the length and depth don't change along the bill.

This time, we'll change the default settings to make the scatter plot look prettier.

Copy-paste the code below in a new R script. Replace the `# insert your code here` comment with your own code.

Add in the `aes()` function (inside `ggplot()`) the arguments `shape = species` and `fill = species` to assign a shape and fill color to your groups, respectively. Finally, add the argument `size = 3` inside `geom_point()` to increase the size of the points.

```{r}
#| label: better-plot
#| eval: false

penguins_colors <- c("#FF9671", "#00D2FC")

# insert your code here
# don't forget to link your code and the lines below with a '+'
  scale_color_manual(values = penguins_colors) + # to change the colors of the points
  scale_shape_manual(values = c(21, 23)) + # to change the shape of the points
  scale_fill_manual(values = alpha(penguins_colors, .4)) + # to add the fill colors and transparency
  labs(
    x = "Body mass (g)",
    y = expression(Bill~area~(mm^{2}))
  ) +
  theme_minimal() + # to change the general theme of the plot
  theme(
    legend.title = element_blank(), # to remove the legend title
    legend.text = element_text(size = 11), # to increase the size of the legend
    axis.text = element_text(size = 11), # to increase the size of the axis tick labels
    axis.title = element_text(size = rel(1.1)) # to increase the size of the axis title, relative to the default size (1.1 times as big as the default size here)
```

Here's the output you're supposed to get:

```{r}
#| label: exercise-output
#| echo: false

library(ggplot2)
penguins_colors <- c("#FF9671", "#00D2FC")

palmerpenguins::penguins |>
  dplyr::filter(species != "Gentoo") |>
  dplyr::mutate(
    bill_area = bill_length_mm * bill_depth_mm
  ) |>
  ggplot(aes(
    x = body_mass_g,
    y = bill_area,
    color = species,
    shape = species,
    fill = species
  )) +
  geom_point(size = 3) +
  scale_color_manual(values = penguins_colors) +
  scale_shape_manual(values = c(21, 23)) +
  scale_fill_manual(values = alpha(penguins_colors, .4)) +
  labs(
    x = "Body mass (g)",
    y = expression(Bill~area~(mm^{2}))
  ) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.text = element_text(size = 11),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = rel(1.1))
  )
```

If you ended up with the same plot, congratulations!

If not, try a bit more. Don't hesitate to go back and read the different sections again. All the information needed to make the plot is in this page. If you have an error, check that you used a proper chaining sign and didn't forget a comma or bracket.
